### 2024-10-07 paper/main.tex
add a first draft of the paper in tex format

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{url}

\title{Three Algorithms for Merging Navigable Graphs}
\author{}
\date{}

\usepackage{natbib}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage[noend]{algpseudocode}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\begin{document}

\maketitle

\section {Local}

% \begin{algorithm}
% \caption{\textsc{Search}($G, q, L$)}\label{alg:search}
% \textbf{Require:} Graph $G = (V, E)$, query $q \in \mathbb{R}^d$, $L \in \mathbb{Z}$ \\
% \textbf{Ensure:} approximate nearest neighbor $v^* \in V$
% \begin{algorithmic}[1]
% \State $C \gets \textsc{InitializeCandidates}()$
% \While{True}
%     \State $u \gets$ nearest unvisited point to $q$ in $C$
%     \State $U \gets \{v \mid (u, v) \in E\}$
%     \For{$v \in U$}
%         \If{$v$ is not visited}
%             \State $C \gets C \cup \{v\}$
%         \EndIf
%     \EndFor
%     \If{$|C| > L$}
%         \State $C \gets$ top $L$ nearest points to $q$ in $C$
%     \EndIf
%     \If{$C$ is not updated}
%         \State \textbf{break}
%     \EndIf
% \EndWhile
% \State \Return nearest point to $q$ in $C$
% \end{algorithmic}
% \end{algorithm}

We start from the basic search.

\begin{algorithm}
\caption{\textsc{LocalSearch}($G, q, C, k, L$)}\label{alg:local_search}
\textbf{Require:} Graph $G = (V, E)$, query $q \in \mathbb{R}^d$, initial set of candidate vertexes $C \subset V$,  $k \in \mathbb{N}$, $L \in \mathbb{N}$ \\
\textbf{Ensure:} approximate k-nearest neighbors $V^* \subset V$
\begin{algorithmic}[1]
% \State $C \gets \textsc{InitializeCandidates}()$
\While{True}
    \State $u \gets$ nearest unvisited point to $q$ in $C$
    \State $U \gets \{v \mid (u, v) \in E\}$
    \For{$v \in U$}
        \If{$v$ is not visited}
            \State $C \gets C \cup \{v\}$
        \EndIf
    \EndFor
    \If{$|C| > L$}
        \State $C \gets$ top $L$ nearest points to $q$ in $C$
    \EndIf
    \If{$C$ is not updated}
        \State \textbf{break}
    \EndIf
\EndWhile
\State \Return nearest point to $q$ in $C$
\end{algorithmic}
\end{algorithm}

$L$ is a parameter that controls the width of search. 
It helps to avoid local minima vertexes. In [HNSW] paper it is named \textbf{ef} parameter (expansion factor).

\begin{algorithm}
\caption{\textsc{HNSW-Search}($\mathcal{H}, q, v_0, k, L, \ell $)}\label{alg:hnsw_search}
\textbf{Require:} The HNSW layers represented as  a sequence of graphs $\mathcal{H} = (G_i)_{i=0}^{l_{max}} )$, query $q \in \mathbb{R}^d$, initial vertex $v_0 \in V$,  $k \in \mathbb{N}$, $L \in \mathbb{N}$, the layer number in which the search should be performed $\ell$ \\    
\textbf{Ensure:} approximate k-nearest neighbors $V^* \subset V$
\begin{algorithmic}[1]
% \State $C \gets \textsc{InitializeCandidates}()$

\State $v^*$ \gets $v_0$ 
\For{$i = l_{max} \; \textbf{to} \; \ell $}
    \State $v^*$ \gets \textsc{LocalSearch}($G=G_i, q=q, C={v^*}, k=1, L=L$)
\EndFor

\State \Return \textsc{LocalSearch}($ G=G_{\ell}, q=q, C={v^*}, k=k, L=L$)
\end{algorithmic}
\end{algorithm}

\section{Merge2}

\begin{algorithm}
\caption{\textsc{HNSW-Merge-Basic}($\mathcal{H}_a, \mathcal{H}_b$)}\label{alg:general_merge}
\textbf{Require:} The HNSW graphs $\mathcal{H}_a = (G^a_i), \mathcal{H}_b = (G^b_i)$ \\
\textbf{Ensure:}  Merged HNSW graphs $\mathcal{H}_c = (G^c_i)$ 
\begin{algorithmic}[1]

% \State $\mathcal{H}_c \gets \emptyset$

\For{$i = 0 \textbf{ to } \max(|\mathcal{H}_a|, |\mathcal{H}_b|) $}
    \State $G^c_i \gets \text{Merge}(G^a_i, G^b_i, i)$
\EndFor

\State $\mathcal{H}_c \gets (G^c_i)_i^{\max(|\mathcal{H}_a|, |\mathcal{H}_b|)}$

\State \Return $\mathcal{H}^c$
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{\textsc{Merge2}($G^a, G^b, \ell, \text{jump\_ef}, \text{local\_ef}, \text{next\_step\_k}, M$)}\label{alg:merge2}
\textbf{Require:} The graphs $G^a=(V^a,E^a), G^b=(V^b,E^b)$ \\
\textbf{Ensure:}  Merged HNSW graphs $\mathcal{H}_c = (G^c_i)$ 
\begin{algorithmic}[1]

\State $E^c \gets \emptyset$
% \State $m \gets \mathcal{H}_a.m_0 \text{ if } \ell = 0 \text{ else } \mathcal{H}_a.m$
\State $\mathcal{V}_{not\_done} \gets V^a \cup V^b$

\While{$\mathcal{V}_{not\_done} \neq \emptyset$}
    \State $v^* \gets \text{random choice from } \mathcal{V}_{not\_done}$

    \State $\mathcal{P}_a  \gets \textsc{HNSW-Search}(\mathcal{H}=\mathcal{H}^a, q=v*, v_0, k, L=\text{jump\_ef}, \ell $)

    \State $\mathcal{P}_b \gets \textsc{HNSW-Search}(\mathcal{H}=\mathcal{H}^b, q=v*, v_0, k, L=\text{jump\_ef}, \ell $)

    \While{True}
        \State $\mathcal{V}_{not\_done} \gets \mathcal{V}_{not\_done} \setminus \{v^*\}$

        \State $\mathcal{C}^a \gets \text{local search in } \mathcal{H}_a.\mathcal{G}[\ell] \text{ from } D_{merged}[v^*]$
        \State $\mathcal{C}^b \gets \text{local search in } \mathcal{H}_b.\mathcal{G}[\ell] \text{ from } D_{merged}[v^*]$

        \If{$v^* \in V^a $}
            \State $\mathcal{C}_{merged} \gets  \{v : (v^*, v) \in E^a \} \cup  \mathcal{C}^b\}$
        \Else
            \State $\mathcal{C}_{merged} \gets  \{v : (v^*, v) \in E^b \} \cup  \mathcal{C}^a\}$
        \EndIf

        % \State $\mathcal{E}_{merged}[v^*] \gets \text{neighborhood\_construction}(\mathcal{C}_{merged}, v^*, m, D_{merged})$

        \State $E^c \gets E^c \cup \{ (v^*, v) : v \in \text{neighborhood\_construction}(\mathcal{C}_{merged}, v^*, m) \}$

        \State $\mathcal{C}^a_{not\_done} \gets \{(v \in \{\mathcal{C}^a_1, \mathcal{C}^a_2, ..., \mathcal{C}^a_{ \text{next\_step\_k} } \} \mid v \in \mathcal{V}_{not\_done} \}$

        \State $\mathcal{C}^b_{not\_done} \gets \{(v \in \{\mathcal{C}^b_1, \mathcal{C}^b_2, ..., \mathcal{C}^b_{ \text{next\_step\_k} } \} \mid v \in \mathcal{V}_{not\_done} \}$

        \State $\mathcal{C}_{not\_done} \gets \mathcal{C}^a_{not\_done} \cup \mathcal{C}^b_{not\_done}$

        \If{$\mathcal{C}_{not\_done} = \emptyset$}
            \State \textbf{break}
        \EndIf

        \State $\text{new\_curr} \gets \arg\min_{(i, d) \in \mathcal{C}^{not\_done}} d$
        \State $v^* \gets \text{new\_curr}$
        \State $\mathcal{P}_a \gets \{i \mid (i, d) \in \mathcal{C}_a\}$
        \State $\mathcal{P}_b \gets \{i \mid (i, d) \in \mathcal{C}_b\}$
    \EndWhile
\EndWhile

\State \Return $E^c$

\end{algorithmic}
\end{algorithm}

\bibliographystyle{plain}
\bibliography{references}
\end{document}

### 2024-10-08 paper/main.tex
finish pseudocode for merge2 algorithm
add  pseudocode for merge1 algorithm

@@
 \usepackage{amsmath}
+\DeclareMathOperator*{\argminA}{arg\,min} % Jan Hlavacek
+\DeclareMathOperator*{\argminB}{argmin}   % Jan Hlavacek
 \usepackage{amsfonts}
@@
 
-\section {Local}
+\section {Search}
 
@@
 
-\State $v^*$ \gets $v_0$ 
+\State $v^* \gets v_0$ 
 \For{$i = l_{max} \; \textbf{to} \; \ell $}
-    \State $v^*$ \gets \textsc{LocalSearch}($G=G_i, q=q, C={v^*}, k=1, L=L$)
+    \State $v^* \gets \textsc{LocalSearch}(G=G_i, q=q, C={v^*}, k=1, L=L)$
 \EndFor
@@
 
-\section{Merge2}
+\section {Neighborhood Construction Strategies}
+
+There are different strategies to form a vertex neighborhood.
+Our proposed merge algorithms are independent in neighborhood procedures.
+Merge algorithm can use a neighborhood construction procedure as an input parameter.
 
 \begin{algorithm}
-\caption{\textsc{HNSW-Merge-Basic}($\mathcal{H}_a, \mathcal{H}_b$)}\label{alg:general_merge}
+\caption{\textsc{KNN-Neighborhood-Construction}$(u,U, k)$}
+\label{alg:knntrategy}
+\textbf{Require:} vertex $u \in V$, neighbor candidates $U \subseteq V$, $k \in \mathbb{N} $\\
+\textbf{Ensure:} selected neighbors $U' \subseteq U$
+\begin{algorithmic}[1]
+
+    \State $U' \gets \text{k-closest neighbors of } U \text{ to } u$
+    \State \Return $U'$
+\end{algorithmic}
+\end{algorithm}
+
+\begin{algorithm}
+\caption{\textsc{RNG-Neighborhood-Construction}$(u, U, m)$}
+\label{alg:rngstrategy}
+\textbf{Require:} vertex $u \in V$, neighbor candidates $U \subseteq V$, maximum neighborhood size $m \in \mathbb{N}$\\
+\textbf{Ensure:} selected neighbors $U' \subseteq U$
+\begin{algorithmic}[1]
+    \State sort $v \in U$ in ascending order of $\delta(u,v)$
+    \State $U' \gets \emptyset$
+    \For{$v \in U$}
+        \State $f \gets \text{true}$
+        \For{$w \in U'$}
+            \If{$\delta(u,v) \geq \delta(v,w)$}
+                \State $f \gets \text{false}$
+                \State \textbf{break}
+            \EndIf
+        \EndFor
+        \If{$f$}
+            \State $U' \gets U' \cup \{v\}$
+        \EndIf
+
+        \If{$|U'| \geq m $}
+            \State \textbf{break}
+        \EndIf
+    \EndFor
+    \State \Return $U'$
+\end{algorithmic}
+\end{algorithm}
+
+\section{Merge Algorithms}
+
+\subsection{HNSW Merge Basic}
+
+\begin{algorithm}
+\caption{\textsc{HNSW-General-Merge}($\mathcal{H}_a, \mathcal{H}_b$)}\label{alg:general_merge}
 \textbf{Require:} The HNSW graphs $\mathcal{H}_a = (G^a_i), \mathcal{H}_b = (G^b_i)$ \\
@@
 
+\subsection{Merge1}
+
 \begin{algorithm}
-\caption{\textsc{Merge2}($G^a, G^b, \ell, \text{jump\_ef}, \text{local\_ef}, \text{next\_step\_k}, M$)}\label{alg:merge2}
-\textbf{Require:} The graphs $G^a=(V^a,E^a), G^b=(V^b,E^b)$ \\
+\caption{\textsc{Merge1}($\mathcal{H}_a, \mathcal{H}_b, \ell, \text{jump\_ef}, \text{local\_ef}, \text{next\_step\_k}, M, m$)}\label{alg:merge1}
+\textbf{Require:} The HNSW graphs $\mathcal{H}_a = (G^a_i), \mathcal{H}_b = (G^b_i)$, the size of the forming neighborhoods $m \in \mathbb{N} $ \\
+\textbf{Ensure:}  Merged HNSW graphs $\mathcal{H}_c = (G^c_i)$ 
+\begin{algorithmic}[1]
+
+\State $E^c \gets \emptyset$
+\State $\mathcal{V}_{not\_done} \gets V^a$
+
+\While{$\mathcal{V}_{not\_done} \neq \emptyset$}
+    \State $v^* \gets \text{random choice from } \mathcal{V}_{not\_done}$
+
+    \State $\mathcal{P}^b  \gets \textsc{HNSW-Search}(\mathcal{H}=\mathcal{H}^b, q=v^*, v_0, k=M, L=\text{jump\_ef}, \ell)$
+
+    \While{True}
+        \State $\mathcal{V}_{not\_done} \gets \mathcal{V}_{not\_done} \setminus \{v^*\}$
+
+        \State $\mathcal{C}^b  \gets \textsc{LocalSearch}(G=G^b, q=v^*, C=\mathcal{P}^b , k=m, L=\text{local\_ef})$
+
+        \State $\mathcal{C} \gets  \{v : (v^*, v) \in E^a \} \cup \mathcal{C}^b$
+
+        \State $E^c[v^*] \gets \textsc{NeighborhoodConstruction}(\mathcal{C}, v^*, m, D_{merged})$
+
+        \State $\mathcal{C}_{not\_done} \gets \{\mathcal{C}^b_1, \mathcal{C}^b_2, ..., \mathcal{C}^b_{ \text{next\_step\_k} } \} \cap \mathcal{V}_{not\_done}$
+
+        \If{$\mathcal{C}_{not\_done} = \emptyset$}
+            \State \textbf{break}
+        \EndIf
+
+        \State $v^* \gets \underset{v \in \mathcal{C}_{not\_done}}{\mathrm{argmin}} \rho(v, v^*)$
+    \EndWhile
+\EndWhile
+
+\State $\mathcal{V}_{not\_done} \gets V^b$
+
+\While{$\mathcal{V}_{not\_done} \neq \emptyset$}
+    \State Repeat the same process for $V^b$ with the roles of $\mathcal{H}_a$ and $\mathcal{H}_b$ swapped.
+\EndWhile
+
+\State \Return $E^c$
+
+\end{algorithmic}
+\end{algorithm}
+
+\subsection{Merge2}
+
+\begin{algorithm}
+\caption{\textsc{Merge2}($G^a, G^b, \ell, \text{jump\_ef}, \text{local\_ef}, \text{next\_step\_k}, M, m$)}\label{alg:merge2}
+\textbf{Require:} The graphs $G^a=(V^a,E^a), G^b=(V^b,E^b)$, the size of the forming neighborhoods $m \in \mathbb{N} $ \\
 \textbf{Ensure:}  Merged HNSW graphs $\mathcal{H}_c = (G^c_i)$ 
@@
 
-    \State $\mathcal{P}_a  \gets \textsc{HNSW-Search}(\mathcal{H}=\mathcal{H}^a, q=v*, v_0, k, L=\text{jump\_ef}, \ell $)
+    \State $\mathcal{P}^a  \gets \textsc{HNSW-Search}(\mathcal{H}=\mathcal{H}^a, q=v^*, v_0, k, L=\text{jump\_ef}, \ell $)
 
-    \State $\mathcal{P}_b \gets \textsc{HNSW-Search}(\mathcal{H}=\mathcal{H}^b, q=v*, v_0, k, L=\text{jump\_ef}, \ell $)
+    \State $\mathcal{P}^b \gets \textsc{HNSW-Search}(\mathcal{H}=\mathcal{H}^b, q=v^*, v_0, k, L=\text{jump\_ef}, \ell $)
 
@@
 
-        \State $\mathcal{C}^a \gets \text{local search in } \mathcal{H}_a.\mathcal{G}[\ell] \text{ from } D_{merged}[v^*]$
-        \State $\mathcal{C}^b \gets \text{local search in } \mathcal{H}_b.\mathcal{G}[\ell] \text{ from } D_{merged}[v^*]$
+        \State $ \mathcal{C}^a  \gets \textsc{LocalSearch}(G=G^a, q=v^*, C=\mathcal{P}^a  , k=m, L=\text{local\_ef})$
+
+        \State $\mathcal{C}^b \gets \textsc{LocalSearch}(G=G^b, q=v^*, C=\mathcal{P}^b , k=m, L=\text{local\_ef})$
 
         \If{$v^* \in V^a $}
-            \State $\mathcal{C}_{merged} \gets  \{v : (v^*, v) \in E^a \} \cup  \mathcal{C}^b\}$
+            \State $\mathcal{C}\gets  \{v : (v^*, v) \in E^a \} \cup  \mathcal{C}^b\}$
         \Else
-            \State $\mathcal{C}_{merged} \gets  \{v : (v^*, v) \in E^b \} \cup  \mathcal{C}^a\}$
+            \State $\mathcal{C} \gets  \{v : (v^*, v) \in E^b \} \cup  \mathcal{C}^a\}$
         \EndIf
@@
 
-        \State $E^c \gets E^c \cup \{ (v^*, v) : v \in \text{neighborhood\_construction}(\mathcal{C}_{merged}, v^*, m) \}$
+        \State $E^c \gets E^c \cup \{ (v^*, v) : v \in \text{neighborhood\_construction}(\mathcal{C}, v^*, m) \}$
 
-        \State $\mathcal{C}^a_{not\_done} \gets \{(v \in \{\mathcal{C}^a_1, \mathcal{C}^a_2, ..., \mathcal{C}^a_{ \text{next\_step\_k} } \} \mid v \in \mathcal{V}_{not\_done} \}$
+        \State $\mathcal{C}^a_{not\_done} \gets \{\mathcal{C}^a_1, \mathcal{C}^a_2, ..., \mathcal{C}^a_{ \text{next\_step\_k} } \} \cap \mathcal{V}_{not\_done}$
 
-        \State $\mathcal{C}^b_{not\_done} \gets \{(v \in \{\mathcal{C}^b_1, \mathcal{C}^b_2, ..., \mathcal{C}^b_{ \text{next\_step\_k} } \} \mid v \in \mathcal{V}_{not\_done} \}$
+        \State $\mathcal{C}^b_{not\_done} \gets \{\mathcal{C}^b_1, \mathcal{C}^b_2, ..., \mathcal{C}^b_{ \text{next\_step\_k} } \}  \cap \mathcal{V}_{not\_done}$
 
@@
 
-        \State $\text{new\_curr} \gets \arg\min_{(i, d) \in \mathcal{C}^{not\_done}} d$
-        \State $v^* \gets \text{new\_curr}$
-        \State $\mathcal{P}_a \gets \{i \mid (i, d) \in \mathcal{C}_a\}$
-        \State $\mathcal{P}_b \gets \{i \mid (i, d) \in \mathcal{C}_b\}$
+        \State $v^* \gets \underset{v \in \mathcal{C}_{not\_done}}{\mathrm{argmin}} \rho(v, v^*)$
+
+        \State $\mathcal{P}_a \gets \mathcal{C}^a$
+        \State $\mathcal{P}_b \gets \mathcal{C}^b$
     \EndWhile
@@
 
+\section{Computational Experiments}
+TODO
+
 \bibliographystyle{plain}

### 2024-10-08 paper/main.tex
fixes in pseudocode of merge algorithms

@@
 \For{$i = 0 \textbf{ to } \max(|\mathcal{H}_a|, |\mathcal{H}_b|) $}
-    \State $G^c_i \gets \text{Merge}(G^a_i, G^b_i, i)$
+    \State $G^c_i \gets \text{Merge}(\mathcal{H}_a, \mathcal{H}_b, \ell=i, \text{algorithm specific params})$
 \EndFor
@@
 \caption{\textsc{Merge1}($\mathcal{H}_a, \mathcal{H}_b, \ell, \text{jump\_ef}, \text{local\_ef}, \text{next\_step\_k}, M, m$)}\label{alg:merge1}
-\textbf{Require:} The HNSW graphs $\mathcal{H}_a = (G^a_i), \mathcal{H}_b = (G^b_i)$, the size of the forming neighborhoods $m \in \mathbb{N} $ \\
-\textbf{Ensure:}  Merged HNSW graphs $\mathcal{H}_c = (G^c_i)$ 
+\textbf{Require:} The HNSW graphs $\mathcal{H}_a = (G^a_i), \mathcal{H}_b = (G^b_i)$, the merging layer number $\ell$, the size of the forming neighborhoods $m \in \mathbb{N}$, parameters $\text{jump\_ef}, \text{local\_ef}, \text{next\_step\_k} \in \mathbb{N}$ \\
+\textbf{Ensure:}  Merged graph $G^c$ 
 \begin{algorithmic}[1]
-
+\State $V^c \gets V^a \cup V^b$
 \State $E^c \gets \emptyset$
+\State $G^a \gets G^a_{\ell} $
+\State $G^b \gets G^b_{\ell} $
 \State $\mathcal{V}_{not\_done} \gets V^a$
@@
 
-        \State $E^c[v^*] \gets \textsc{NeighborhoodConstruction}(\mathcal{C}, v^*, m, D_{merged})$
+        \State $E^c \gets E^c \cup  \{ (v^*, v)  : v \in \textsc{NeighborhoodConstruction}(\mathcal{C}, v^*, m) \}$
 
@@
 
-\State \Return $E^c$
+\State \Return $G^c=(V^c,E^c)$
 
@@
 \begin{algorithm}
-\caption{\textsc{Merge2}($G^a, G^b, \ell, \text{jump\_ef}, \text{local\_ef}, \text{next\_step\_k}, M, m$)}\label{alg:merge2}
-\textbf{Require:} The graphs $G^a=(V^a,E^a), G^b=(V^b,E^b)$, the size of the forming neighborhoods $m \in \mathbb{N} $ \\
-\textbf{Ensure:}  Merged HNSW graphs $\mathcal{H}_c = (G^c_i)$ 
+\caption{\textsc{Merge2}($\mathcal{H}^a, \mathcal{H}^b,  \ell, \text{jump\_ef}, \text{local\_ef}, \text{next\_step\_k}, M, m$)}\label{alg:merge2}
+\textbf{Require:} The HNSW graphs $\mathcal{H}_a = (G^a_i), \mathcal{H}_b = (G^b_i)$, the number of merged layers, the size of the forming neighborhoods $m \in \mathbb{N} $, parameters $\text{jump\_ef}, \text{local\_ef}, \text{next\_step\_k}, M \in \mathbb{N}$ \\
+\textbf{Ensure:}  Merged graph $G^c$ 
 \begin{algorithmic}[1]
 
+\State $V^c \gets V^a \cup V^b$
 \State $E^c \gets \emptyset$
+\State $G^a \gets G^a_{\ell} $
+\State $G^b \gets G^b_{\ell} $
 % \State $m \gets \mathcal{H}_a.m_0 \text{ if } \ell = 0 \text{ else } \mathcal{H}_a.m$
@@
 
-        % \State $\mathcal{E}_{merged}[v^*] \gets \text{neighborhood\_construction}(\mathcal{C}_{merged}, v^*, m, D_{merged})$
-
         \State $E^c \gets E^c \cup \{ (v^*, v) : v \in \text{neighborhood\_construction}(\mathcal{C}, v^*, m) \}$

### 2024-10-08 paper/main.tex
add pseudocode for naive-merge alg
add text description of general, naive, and merge1 algorihtms

@@
 \usepackage{amssymb}
+\usepackage{nameref}
 \begin{document}
@@
 
-We start from the basic search.
+We start from the basic search over the graph.
 
@@
 
-\subsection{HNSW Merge Basic}
+We provide 3 merge algorithms: "Naive Merge", "Merge1", and "Merge2".
+We also refer to these algorithm as layer-merge algorithms.
+"Naive Merge" is a straight forward method, while "Merge1" and "Merge2" are more advanced.
+Together, these two algorithms constitute the main contribution of the work.
+The HNSW structure is a sequence of graphs, called layers.
+We start from description of a general algorithm for merging multi layers structure.
+
+\subsection{HNSW General Merge}
+
+HNSW-General-Merge algorithm \ref{alg:general_merge} is a straight forward.
+It uses one of the layer-merge algorithm to merge each HNSW layer.
+To able layer-merge algorithm to perform fast search using multi-layer structure of HNSW, HNSW-General-Merge algorithm passes to layer merge algorithm both HNSW structures $\mathcal{H}_a, \mathcal{H}_b$, and the layer number $\ell=i$ to which the merge should be performed.
+An order of merging layers doesn't mater.
+Thus merging of layers can be done in parallel.
 
@@
 \begin{algorithmic}[1]
-
-% \State $\mathcal{H}_c \gets \emptyset$
 
@@
 
+\subsection{Naive Merge}
+The Naive Merge algorithm \ref{alg:merge_naive} surprisingly is very simple.
+It takes one vertex $v^*$ by one from the graph $G^a$.
+Then algorithm prepares the candidate set $\mathcal{C}$ for a new neighborhood (lines 7,8).
+It uses \textsc{HNSW-Search} to get $m$-closest vertices to $v^*$ in the graph $G^b$, and combines with an old neighborhood of the vertex $v^*$.
+After that in the line 9 the candidate set $\mathcal{C}$ is passed into the neighborhood construction function, which selects the vertices for the new merged neighborhood for $v^*$.
+Lines 10-13 contains the same logic for the forming new neigborhoods of graph $G^b$
+
+\begin{algorithm}
+\caption{\textsc{Merge-Naive}($\mathcal{H}_a, \mathcal{H}_b,  \ell, \text{search\_ef}$)}\label{alg:merge_naive}
+\textbf{Require:} The HNSW graphs $\mathcal{H}_a = (G^a_i)$, $\mathcal{H}_b = (G^b_i)$, and the level number $\ell$, search parameter $\text{search\_ef}$ \\
+\textbf{Ensure:} Merged graph $G^c$ 
+\begin{algorithmic}[1]
+\State $V^c \gets V^a \cup V^b$
+\State $E^c \gets \emptyset$
+\State $G^a \gets G^a_{\ell}$
+\State $G^b \gets G^b_{\ell}$
+\State $m \gets \mathcal{H}_a.m_0$ if $\ell = 0$ else $\mathcal{H}_a.m$
+
+\For{$v^* \in V^a$}
+    \State $\mathcal{C}^b \gets \textsc{HNSW-Search}(\mathcal{H} = \mathcal{H}^b, q = v^*, v_0, k = m, L = \text{search\_ef}, \ell)$
+    \State $\mathcal{C} \gets \{v : (v^*, v) \in E^a \} \cup \mathcal{C}^b$
+    \State $E^c \gets E^c \cup \{(v^*, v) : v \in \textsc{neighborhood\_construction}(\mathcal{C}, v^*, m)\}$
+\EndFor
+
+\For{$v^* \in V^b$}
+    \State $\mathcal{C}^a \gets \textsc{HNSW-Search}(\mathcal{H} = \mathcal{H}^a, q = v^*, v_0, k = m, L = \text{search\_ef}, \ell)$
+    \State $\mathcal{C} \gets \{v : (v^*, v) \in E^b \} \cup \mathcal{C}^a$
+    \State $E^c \gets E^c \cup \{(v^*, v) : v \in \textsc{neighborhood\_construction}(\mathcal{C}, v^*, m)\}$
+\EndFor
+
+\State \Return $G^c = (V^c, E^c)$
+\end{algorithmic}
+\end{algorithm}
+
 \subsection{Merge1}
+The most effort of \textsc{Merge-Naive} algorithm lies in the set of neighborhood candidates from the opposite graph utilising \textsc{HNSW-Search} procedure, that traverses layers graph from the top-level down to layer number $\ell$ every time.
+The number of computations can be reduced if we select the next vertex $v^*$ close to the previous one (line 15), instead of choosing it randomly.
+Thus, for new $v^*$ the neighborhood candidates will be  also close to the previous candidates set.
+To search this new neighborhood candidates we can use \textsc{LocalSearch} procedure, that traversing the same graph staring from the previous neighborhood candidates set $\mathcal{P}^b$ (line 11,14).
+In the line 14 in set $\mathcal{P}^b$ we keep only $M$-closest to $v^*$ candidates.\\
+In the line 15 almost at each iteration we try select new $v^*$ vertex close to the previous  $v^*$ that was not already processed.  
+To ensure that new $v^*$ vertex are not very far from the previous $v^*$ we bound the size of results of \textsc{LocalSearch} procedure controlling by "next\_step\_k" parameter.
+Once \textsc{LocalSearch} can't find enough close not processed vertex, we choice new $v^*$ from the set $\mathcal{V}_not\_done$ randomly.
 
@@
 
-        \State $\mathcal{C}_{not\_done} \gets \{\mathcal{C}^b_1, \mathcal{C}^b_2, ..., \mathcal{C}^b_{ \text{next\_step\_k} } \} \cap \mathcal{V}_{not\_done}$
+        \State $\mathcal{P}^b \gets \{\mathcal{C}^b_1, \mathcal{C}^b_2, ..., \mathcal{C}^b_M \} $
 
-        \If{$\mathcal{C}_{not\_done} = \emptyset$}
+        \State $\mathcal{C}^a  \gets \textsc{LocalSearch}(G=G^a, q=v^*, C=\{v^*\} , k=\text{next\_step\_k}, L=\text{next\_step\_ef})$
+
+        \State $\mathcal{C}^a \gets \mathcal{C}^a \cap \mathcal{V}_{not\_done}$
+
+        \If{$\mathcal{C}^a = \emptyset$}
             \State \textbf{break}
@@
 
-        \State $v^* \gets \underset{v \in \mathcal{C}_{not\_done}}{\mathrm{argmin}} \rho(v, v^*)$
+        \State $v^* \gets \mathcal{C}^a_1$
     \EndWhile
@@
 \While{$\mathcal{V}_{not\_done} \neq \emptyset$}
-    \State $v^* \gets \text{random choice from } \mathcal{V}_{not\_done}$
+    \State $v^* \gets \text{randomly choice from } \mathcal{V}_{not\_done}$
 

### 2024-10-08 paper/main.tex
add text description of merge2 algorithm

@@
 It takes one vertex $v^*$ by one from the graph $G^a$.
-Then algorithm prepares the candidate set $\mathcal{C}$ for a new neighborhood (lines 7,8).
+Then algorithm prepares the candidate set $\mathcal{C}$ for a new neighborhood (lines 6,7).
 It uses \textsc{HNSW-Search} to get $m$-closest vertices to $v^*$ in the graph $G^b$, and combines with an old neighborhood of the vertex $v^*$.
-After that in the line 9 the candidate set $\mathcal{C}$ is passed into the neighborhood construction function, which selects the vertices for the new merged neighborhood for $v^*$.
-Lines 10-13 contains the same logic for the forming new neigborhoods of graph $G^b$
+After that in the line 8 the candidate set $\mathcal{C}$ is passed into the neighborhood construction function, which selects the vertices for the new merged neighborhood for $v^*$.
+Lines 9-12 contains the same logic for the forming new neighborhoods of graph $G^b$
 
@@
 \State $G^b \gets G^b_{\ell}$
-\State $m \gets \mathcal{H}_a.m_0$ if $\ell = 0$ else $\mathcal{H}_a.m$
+% \State $m \gets \mathcal{H}_a.m_0$ if $\ell = 0$ else $\mathcal{H}_a.m$
 
@@
 To ensure that new $v^*$ vertex are not very far from the previous $v^*$ we bound the size of results of \textsc{LocalSearch} procedure controlling by "next\_step\_k" parameter.
-Once \textsc{LocalSearch} can't find enough close not processed vertex, we choice new $v^*$ from the set $\mathcal{V}_not\_done$ randomly.
+Once \textsc{LocalSearch} can't find enough close not processed vertex, in line 7 we choice new $v^*$ from the set $\mathcal{V}_{not\_done}$ randomly.
+\\
+After we have processed all vertices from the the graph $G^a$. We do the same for vertices of graph $G^b$ (lines 21-22).
 
@@
 \subsection{Merge2}
+
+\textsc{Merge2} algorithm is similar to \textsc{Merge1} utilise \textsc{LocalSearch} procedure to reduce computation effort.
+The difference is that \textsc{Merge1} algorithm the next $v^*$ vertex choice from the same graph, while \textsc{Merge2} looks for new vertex $v^*$ in the both graphs $G^a$, and $G^b$.
+Thus in line 24  $v^*$  is chosen from the set $\mathcal{C}_{not\_done}$, which of not processed vertex from the both graph (line 21).
+The intuition laying behind \textsc{Merge2} algorithm is that allowing to choice new $v^*$ from the both graphs, we reduce the number of times what $v^*$ is chosen randomly, thereby minimise the number of using more expensive search procedure \textsc{HNSW-Search}.
 

### 2024-10-08 paper/main.tex
figures to computation results section

@@
 \usepackage{graphicx}
+\usepackage{subcaption}
 \usepackage{algorithm}
@@
 \section{Computational Experiments}
-TODO
+We implemented proposed algorithms using python.
+In order to be independent of the implementation details and low level optimisation, we do comparison of algorithms based on the number of distance computation during the merge process.
+
+\begin{figure}
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=1.\linewidth]{figs/merge-effort-comparison.pdf}
+  \caption{Merging efforts}
+  \label{fig:sfig1}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[width=1.\linewidth]{figs/recall-comparsion.pdf}
+  \caption{Recall}
+  \label{fig:sfig2}
+\end{subfigure}
+\caption{Merging algorithm comparison}
+\label{fig:fig}
+\end{figure}
 

### 2024-10-08 paper/main.tex
Extend computation results section

@@
 We implemented proposed algorithms using python.
-In order to be independent of the implementation details and low level optimisation, we do comparison of algorithms based on the number of distance computation during the merge process.
+For experiments we use Sift1m dataset divided into two subset of 500k vectors at each.
+Our algorithm merge that two subsets of 500k vectors.
+In order to be independent of the implementation details and low level optimisation, we do comparison of the algorithms based on the number of distance computation during the merge process.
+\ref{fig:sfig1} shows that \textsc{Merge1} about 5 times less computations than naive algorithm.
+Surprisingly \textsc{Merge2} algorithm do slightly more computations than \textsc{Merge1}, nevertheless it significantly outperform \textsc{merge-naive} algorithm.\\
+The other important characteristics of merging process is how accurate the merge is done.
+It can be easily verified for the case of \textsc{KNN-Neighborhood-Construction} procedure, by comparing the set of merged neighborhoods with k-nearest-neighbor set for each vertex.
+However to follow the same way in the case of \textsc{RNG-Neighborhood-Construction} is not good idea, because the \textsc{RNG-Neighborhood-Construction} is one of the possible heuristic for Voronoi-cell approximation.
+Therefore we measure the recall-distance computation trade off of \textsc{HNSW-Search} procedure on the HNSW graphs before merging, and after.
+As can be seen from the \ref{fig:sfig2} the recall/computational trade off of naive algorithm is slightly better than  \textsc{Merge1}, and  \textsc{Merge2} has.
 
@@
   \includegraphics[width=1.\linewidth]{figs/recall-comparsion.pdf}
-  \caption{Recall}
+  \caption{Recall/computation trade off}
   \label{fig:sfig2}

### 2024-10-09 paper/main.tex
insert introduction section with the help of gpt

@@
 
+\section{Introduction}
+In this paper we presented three algorithms for merging navigable graphs, specifically Hierarchical Navigable Small World (HNSW) graphs, which are used for efficient nearest neighbor searches.
+
+Merging HNSW graphs is important for optimizing vector search systems, especially in large-scale databases and distributed environments.
+It consolidates data from different partitions or batches, improving search efficiency and unifying querying processes.
+For vector index compaction, merging reduces redundancy, saves memory, and improves search performance by compacting the graph structure.
+It supports incremental updates, allowing new data to be added without rebuilding the entire index.
+This is crucial for dynamic systems where data grows or changes frequently.
+Merging also helps combine embeddings from different models or datasets into a unified index, benefiting applications like search engines, recommendation systems, and machine learning pipelines.
+It provides a single query interface for cross-organizational or distributed data sources, ensuring seamless access.
+In recommender systems, it enhances performance and accuracy by unifying item indices.
+Merged graphs lead to more efficient memory usage and faster query response times.
+It enables better management of vector embeddings in vector databases, minimizing complexity.
+The process is especially useful for handling incremental updates, reducing the need for full index rebuilds.
+Ultimately, merging HNSW graphs boosts performance in vector search and retrieval tasks across diverse applications.
+
+The three algorithms proposed are “Naive Merge”, “Merge1”, and “Merge2”, each progressively building on the previous to offer more optimized and efficient strategies for merging two graph structures.
+
+The algorithms are designed to address the problem of merging two distinct HNSW structures, $\mathcal{H}_a$ and $\mathcal{H}_b$, into a new, merged structure $\mathcal{H}_c$.
+This process is particularly useful in dynamic scenarios where multiple HNSW graphs may need to be combined without fully recomputing from scratch
+
 \section {Search}
-
-% \begin{algorithm}
-% \caption{\textsc{Search}($G, q, L$)}\label{alg:search}
-% \textbf{Require:} Graph $G = (V, E)$, query $q \in \mathbb{R}^d$, $L \in \mathbb{Z}$ \\
-% \textbf{Ensure:} approximate nearest neighbor $v^* \in V$
-% \begin{algorithmic}[1]
-% \State $C \gets \textsc{InitializeCandidates}()$
-% \While{True}
-%     \State $u \gets$ nearest unvisited point to $q$ in $C$
-%     \State $U \gets \{v \mid (u, v) \in E\}$
-%     \For{$v \in U$}
-%         \If{$v$ is not visited}
-%             \State $C \gets C \cup \{v\}$
-%         \EndIf
-%     \EndFor
-%     \If{$|C| > L$}
-%         \State $C \gets$ top $L$ nearest points to $q$ in $C$
-%     \EndIf
-%     \If{$C$ is not updated}
-%         \State \textbf{break}
-%     \EndIf
-% \EndWhile
-% \State \Return nearest point to $q$ in $C$
-% \end{algorithmic}
-% \end{algorithm}
 

### 2024-10-09 merge_hnsw.py
In merge1 use parameter M to get set of new starting points

@@ -6,15 +6,11 @@ import random
 
 def hnsw_general_merge(hnsw_a, hnsw_b, merged_data, layer_merge_func):
     hnsw_merged = HNSW(distance_func=hnsw_a.distance_func, m=hnsw_a.m, m0=hnsw_a.m0, ef=hnsw_a.ef, ef_construction=hnsw_a.ef_construction, neighborhood_construction = hnsw_a.neighborhood_construction)
-    hnsw_merged.data = merged_data
-    hnsw_merged.graphs = [] # sequence of merged graphs (layers)
+    hnsw_merged.data = merged_data 
+    hnsw_merged.graphs = [] # sequence of merged graphs (layers) denoted as $G^C_i$
     levels_merged_max = max(len(hnsw_a.graphs), len(hnsw_b.graphs))
     levels_merged_min = min(len(hnsw_a.graphs), len(hnsw_b.graphs))
-
-    if len(hnsw_a.graphs) >= len(hnsw_b.graphs):
-        hnsw_merged.enter_point = hnsw_a.enter_point
-    else:
-        hnsw_merged.enter_point = hnsw_b.enter_point
+    
 
     for level in range(levels_merged_min): 
         print('Merging level:', level)
@@ -25,6 +21,11 @@ def hnsw_general_merge(hnsw_a, hnsw_b, merged_data, layer_merge_func):
             hnsw_merged.graphs.append(hnsw_a.graphs[level])
         else:
             hnsw_merged.graphs.append(hnsw_b.graphs[level])
+
+    if len(hnsw_a.graphs) >= len(hnsw_b.graphs):
+        hnsw_merged.enter_point = hnsw_a.enter_point
+    else:
+        hnsw_merged.enter_point = hnsw_b.enter_point
 
     return hnsw_merged
 
@@ -75,7 +76,7 @@ def merge1_layer(hnsw_a, hnsw_b, merged_data, level, jump_ef, local_ef, next_ste
     next_step_ef – a purpose of this parameter is similar {next_step_k}
     M            – number of starting point for the local search.  
     '''
-    merged_edges = {}
+    merged_edges = {} # merged graph
     not_done = set(hnsw_a.graphs[level].keys())
     m = hnsw_a.m0 if level == 0 else hnsw_a.m
     
@@ -98,7 +99,8 @@ def merge1_layer(hnsw_a, hnsw_b, merged_data, level, jump_ef, local_ef, next_ste
[…]

### 2024-10-09 paper/main.tex
Rewrite introdution section

--- 2024-10-08 paper/main.tex
@@
 
+\section{Introduction}
+In recent years, navigable small-world graphs have become a cornerstone of efficient approximate nearest neighbor (ANN) search in high-dimensional spaces, a problem prevalent in fields such as recommendation systems, image retrieval, and natural language processing.
+Algorithms like Hierarchical Navigable Small World (HNSW) and its predecessor Navigable Small World (NSW) provide a robust and scalable framework for organizing data into graph structures that support fast similarity searches.
+However, one limitation inherent to these methods is the lack of a natural mechanism for node deletion, making the need for merging and compacting graphs crucial for maintaining the efficiency and accuracy of the search index over time.
+
+The HNSW algorithm, widely adopted for its performance and scalability, supports dynamic insertions but lacks native support for node deletion.
+As databases grow and evolve, some nodes become outdated or irrelevant, requiring a process to filter out these deleted vertices.
+Merging operations allow for a form of "data compaction," where the graphs from two or more databases are merged, and nodes marked for deletion are omitted from the final structure.
+This process not only improves space efficiency but also enhances the accuracy of future searches by eliminating superfluous data points.
+
+In this work, we propose three novel algorithms designed to efficiently merge navigable graphs, focusing on HNSW but extending applicability to other graph-based data structures like NSW, NSG.
+The merging algorithms—"Naive Merge," "Merge1," and "Merge2"—offer progressively advanced methods to combine graphs while preserving the properties that make navigable graphs effective for nearest neighbor search.
+These methods ensure that the resulting merged index excludes deleted nodes and maintains the structural integrity needed for fast search operations.
+
 \section {Search}
-
-% \begin{algorithm}
-% \caption{\textsc{Search}($G, q, L$)}\label{alg:search}
-% \textbf{Require:} Graph $G = (V, E)$, query $q \in \mathbb{R}^d$, $L \in \mathbb{Z}$ \\
-% \textbf{Ensure:} approximate nearest neighbor $v^* \in V$
-% \begin{algorithmic}[1]
-% \State $C \gets \textsc{InitializeCandidates}()$
-% \While{True}
-%     \State $u \gets$ nearest unvisited point to $q$ in $C$
-%     \State $U \gets \{v \mid (u, v) \in E\}$
-%     \For{$v \in U$}
-%         \If{$v$ is not visited}
-%             \State $C \gets C \cup \{v\}$
-%         \EndIf
-%     \EndFor
-%     \If{$|C| > L$}
-%         \State $C \gets$ top $L$ nearest points to $q$ in $C$
-%     \EndIf
-%     \If{$C$ is not updated}
-%         \State \textbf{break}
-%     \EndIf
-% \EndWhile
-% \State \Return nearest point to $q$ in $C$
-% \end{algorithmic}
-% \end{algorithm}
 

### 2024-10-09 paper/main.tex
Rename parameters in pseudocode of neighborhood construction procedure
add several citations to the paper

@@
 \usepackage{url}
+\usepackage{authblk}
 
 \title{Three Algorithms for Merging Navigable Graphs}
-\author{}
+\author[1]{Alexander Ponomarenko}
 \date{}
+
+\affil[1]{YugaByte LLC}
+% \affil[2]{HSE Univerity}
 
@@
 In recent years, navigable small-world graphs have become a cornerstone of efficient approximate nearest neighbor (ANN) search in high-dimensional spaces, a problem prevalent in fields such as recommendation systems, image retrieval, and natural language processing.
-Algorithms like Hierarchical Navigable Small World (HNSW) and its predecessor Navigable Small World (NSW) provide a robust and scalable framework for organizing data into graph structures that support fast similarity searches.
+Algorithms like Hierarchical Navigable Small World (HNSW) \cite{hnsw} and its predecessor Navigable Small World (NSW) \cite{nsw2011,nsw2012,nsw2014} provide a robust and scalable framework for organizing data into graph structures that support fast similarity searches.
 However, one limitation inherent to these methods is the lack of a natural mechanism for node deletion, making the need for merging and compacting graphs crucial for maintaining the efficiency and accuracy of the search index over time.
@@
 
-In this work, we propose three novel algorithms designed to efficiently merge navigable graphs, focusing on HNSW but extending applicability to other graph-based data structures like NSW, NSG.
+In this work, we propose three novel algorithms designed to efficiently merge navigable graphs, focusing on HNSW but extending applicability to other graph-based data structures like NSW \cite{nsw2011}, NSG \cite{NSG}.
 The merging algorithms—"Naive Merge," "Merge1," and "Merge2"—offer progressively advanced methods to combine graphs while preserving the properties that make navigable graphs effective for nearest neighbor search.
@@
 
-We start from the basic search over the graph.
+We start from the basic search over the sigle graph.
 
@@
 \begin{algorithm}
-\caption{\textsc{KNN-Neighborhood-Construction}$(u,U, k)$}
+\caption{\textsc{KNN-Neighborhood-Construction}$(v^*,C, k)$}
 \label{alg:knntrategy}
-\textbf{Require:} vertex $u \in V$, neighbor candidates $U \subseteq V$, $k \in \mathbb{N} $\\
-\textbf{Ensure:} selected neighbors $U' \subseteq U$
+\textbf{Require:} vertex $v^* \in V$, neighbor candidates $C \subseteq V$, $k \in \mathbb{N} $\\
+\textbf{Ensure:} selected neighbors $C' \subseteq C$
 \begin{algorithmic}[1]
 
-    \State $U' \gets \text{k-closest neighbors of } U \text{ to } u$
-    \State \Return $U'$
+    \State $C' \gets \text{k-closest neighbors of } C \text{ to } v^*$
+    \State \Return $C'$
 \end{algorithmic}
@@
 \begin{algorithm}
-\caption{\textsc{RNG-Neighborhood-Construction}$(u, U, m)$}
+\caption{\textsc{RNG-Neighborhood-Construction}$(v^*, C, m)$}
 \label{alg:rngstrategy}
-\textbf{Require:} vertex $u \in V$, neighbor candidates $U \subseteq V$, maximum neighborhood size $m \in \mathbb{N}$\\
-\textbf{Ensure:} selected neighbors $U' \subseteq U$
+\textbf{Require:} vertex $v^* \in V$, neighbor candidates $C \subseteq V$, maximum neighborhood size $m \in \mathbb{N}$\\
+\textbf{Ensure:} selected neighbors $C' \subseteq C$
 \begin{algorithmic}[1]
-    \State sort $v \in U$ in ascending order of $\delta(u,v)$
-    \State $U' \gets \emptyset$
+    \State sort $v \in C$ in ascending order of $\rho(v^*,v)$
+    \State $C' \gets \emptyset$
     \For{$v \in U$}
         \State $f \gets \text{true}$
-        \For{$w \in U'$}
-            \If{$\delta(u,v) \geq \delta(v,w)$}
+        \For{$w \in C'$}
+            \If{$\delta(v^*,v) \geq \delta(v,w)$}
                 \State $f \gets \text{false}$
@@
         \If{$f$}
-            \State $U' \gets U' \cup \{v\}$
+            \State $C' \gets C' \cup \{v\}$
         \EndIf
 
-        \If{$|U'| \geq m $}
+        \If{$|C'| \geq m $}
             \State \textbf{break}
@@
     \EndFor
-    \State \Return $U'$
+    \State \Return $C'$
 \end{algorithmic}
